Type Visualization for R

### 2020-06-22 paper/main.tex
paper: More edits

\documentclass{vgtc}                          % final (conference style)
%\documentclass[review]{vgtc}                 % review
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[electronic]{vgtc}             % electronic version

@@

\usepackage{caption}
\captionsetup[table]{skip=10pt}
\usepackage{tabularx}
\usepackage{todonotes} %TODO: Remove after all TODO's are gone

%% ===================================== Commands
\newcommand{\code}[1]{\texttt{\small{#1}}}
\newcommand{\typevis}{{\sc TypeVis}\xspace}

@@

\typevis can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases,
or weed out language designs that are incompatible with existing programs.
\typevis was built in support of a new gradual type system for R,
whose adoption depends on integrating well with existing code.
However, the visualization is not limited to R and can be used to
analyze type usage in any language where similar data is available.

@@

Being a dynamically-typed language
with widespread use in a single domain,
R is a good target for a gradual type system.
A gradual type system permits the integration of
static and dynamic typing~\cite{siek:2006,tobin-hochstadt:2006}.
Such a type system is capable of seamlessly mixing typed and untyped regions of code.

For statically-typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the static type system precedes any code written
in the language.
Gradual type systems, by contrast, are often designed atop
an existing dynamically-typed language~\cite{tobin-hochstadt:2016}.
Many R programs already exist,
so a successful gradual type system must
work well with extant code.

We are not aware of any previous work that applies
visualization techniques for the purpose of programming language design.
However, there are existing publications on understanding
software systems through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces
to improve developer comprehension of large software systems.
Their tool applies pruning and summarization techniques to reduce trace size.
They focus on call topology,
describing the structure of function invocations,
rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
\typevis has elements of both,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information,
allowing the user to gain further understanding of the system
under examination.
This includes grouping methods that come from the same class---a
basic form of type data.

@@

The data set we used comes from
Turcotte et al.~\cite{turcotte:2020}
and consists of traces from 412 packages,
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN are curated and satisfy
certain quality standards.
Out of the over $15,000$ packages
distributed on CRAN,
the subset of 412 packages considered here
all had code coverage above 65\%.
\todo{Confirm with Alexi this paragraph. Get proper BibTeX entry.}

Execution traces were recorded by running a dynamic analysis
over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of function type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, the function call {\tt !(FALSE)}
would be recorded as the type signature {\tt logical $\to$ logical},
meaning that the function {\tt !} took a logical value (i.e. a boolean)
and returned a logical value.
The type tags are assigned by the dynamic analysis
and refine the actual type tags in R.
For instance, the analysis
assigns the scalar tag {\tt double} to a singleton vector of double values
because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \typevis was designed for the analysis of types,
not all information in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

@@

We interviewed the primary researcher developing a gradual type system for R
to inventory relevant domain-specific tasks.
Additionally, we sought feedback throughout the creation of \typevis
to ensure it was adequately meeting requirements.
The domain-specific goals identified were:

@@

\typevis is an interactive web-based visualization
that permits real-time exploration,
even with large data sets.
The default data set,
described in \S\ref{sec:data},
contains over a million rows.
Figure~\ref{fig:typevis} shows
the heart of \typevis
in its initial configuration.
Most aspects of the system can be tweaked
by the user if desired.

@@

The right-hand panel of figure~\ref{fig:typevis}
contains three elements: a search bar,
a package treemap,
and a function treemap.
All three are linked.
Selecting a package
in the top treemap will cause the
bottom view to show only functions available in that package.
Entering a function name in the search bar
will select it in the function treemap.

These different components support different modes of search.
If a user has a specific function in mind,
the search bar supports autocomplete
that can select one function among thousands.
However, if one wants to browse
all available functions and packages,
the treemap view is more applicable.
Area in the bottom treemap conveys the log-scaled
frequency of function calls,
and correspondingly the top treemap
is scaled based on invocations
of functions defined in that package.
Log-scaling and pagination keep
elements of the treemap legible,
even when absolute size differences are significant.

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
The packages in our data are not deeply structured,
so the treemaps in figure~\ref{fig:typevis} are only one level deep.
However, if the data set recorded deeply nested packages,
\typevis could be modified to handle that.

If configured, \typevis allows for multiple function selection.
In this example, the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are selected.
These functions are rendered simultaneously in the type flow
panel, allowing the user to make comparisons between them.

@@

Prominently featured in the center of figure~\ref{fig:typevis}
is a parallel sets diagram encoding the type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara et al.~\cite{kosara:2006},
are ideally suited for type signatures.
Each data point can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n)$
corresponding to the function name,
argument types,
and return type.
Every component of that tuple is categorical,
and we are interested in visualizing the frequencies of these points.
This is the exact use-case for parallel sets.

Flows begin at a node labeling a function,
curve across each argument type in order,
and terminate at the return type.
Widths are proportional to how many times a function
is called with that signature during analysis.
Flow segments are filled with a hue-varying color scale,
determined by the type where the segment ends.
Beneath each node label is an approximate count of either:
how many times the function was invoked if the node represents a function,
or how often a value of that type was recorded during analysis if the node represents a tuple.

In figure~\ref{fig:typevis},
the {\tt length.unit} function is called with
two different type signatures at run-time.
Most frequently, the function is used with the
signature {\tt double $\to$ integer},
but a substantial fraction of the time it is
used as {\tt double[] $\to$ integer}.
Here, {\tt double[]} stands for an array of
doubles with arbitrary dimension.

Interactivity significantly enhances the
usability of the parallel sets diagram.
Hovering over a flow will highlight the
entire path of the flow
and will fade out all other functions.
When displaying many flows at the same time,
highlighting becomes critical
for user comprehension.
Additional quantitative information
about the flow is supplied on-demand while hovering.
Clicking a flow will focus on that function
only, filtering away all others.

Flow layout is a necessary consideration
for a legible visualization~\cite{sugiyama:1981}.
Figure~\ref{fig:decross} exemplifies what
happens if flow layout is not adequately computed.
The top layout was generated without any
flow decrossing algorithm,
while the bottom layout applies optimal decrossing.
Even with highlighting,
the visualization is almost impossible to understand
due to excessive flow overlap.
To maintain real-time performance,
we use two different decrossing methods.
First, the system attempts to construct
a layout with the globally optimal solution,
minimizing the amount of flow crossing by
solving a mixed-integer linear program.
Unfortunately, for even moderately
sized graphs this can take too long to compute.
Therefore, after about $1$ second,
the optimal algorithm is terminated
if it has not finished, and a faster
algorithm that only minimizes local crossing between layers
computes the layout.
While not ideal, it is much better than no decrossing
at all.

@@

Since the amount of data a user may query is large,
the application consists of both a frontend and a backend.
This avoids excessive memory consumption and speeds up computations,
as the client queries data from the server on an need-to-know basis.

@@

\typevis suffers from some shortcomings
and does not fully address the entire range of tasks
a language designer may need to perform.

In particular, R supports a range of data types
including ones containing dimensionality information.
For example, an R array may have type {\tt integer[3]}
indicating that it is an array of length $3$.
This will get compressed down into a single array of
type {\tt integer[]}.
Reducing the number of types can make the visualization
more understandable, at the expensive of losing
precision in type information.
Ideally a user would be able to interactively tune
the level of granularity based on their interest.
Such a feature would be especially helpful for R
since values can come attached with lots of relevant
data such as the dynamic dispatch method
or user-specified metadata.

@@

We have presented \typevis, an interactive visualization
of run-time type information
that supports real-time analysis and exploration.
The tool was developed for a massive data set of
execution traces from a corpus of popular R packages.

\typevis provides facilities for filtering, searching,
understanding, and comparing type information.
Signatures are represented as flows traversing
each argument type and return type in sequence.
Interactions provide additional details and
assist users with comprehension and navigation.

Our system is aimed at elucidating how R programmers
use the language in the field.
Specifically, the data and visualization will be used to
inform the design of a future gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any dynamically-typed language.
The visualization could yield insights
useful for purposes other than gradual typing,
and beyond what we have considered.

We hope that, in the future, programming language designers
will collect and use empirical data about their languages.
Such data could be used to make informed decisions about design,
where the cost of making a mistake is high.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
\typevis is a first step in this direction.

>>>>

%\documentclass{vgtc}                          % final (conference style)
\documentclass[review]{vgtc}                 % review
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[electronic]{vgtc}             % electronic version

@@

\usepackage{caption}
\captionsetup[table]{skip=10pt}
\usepackage{tabularx}
\usepackage{todonotes} %TODO: Remove after all TODO's are gone
\usepackage{lineno}
\linenumbers

%% ===================================== Commands
\newcommand{\code}[1]{\texttt{\small{#1}}}
\newcommand{\typevis}{{\sc TypeVis}\xspace}
\newcommand{\figref}[1]{Figure~\ref{#1}\xspace}
\newcommand{\pkgNumAnalyzed}{412\xspace}
\newcommand{\settingsTab}{{\sc Settings}\xspace}

@@

\typevis can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases,
or weed out language designs that are incompatible with existing programs.
\typevis was built in support of a new gradual type system for R,
whose adoption depends on integrating well with existing code.
However, the visualization is not limited to R and can be used to
analyze type usage in any language where similar data are available.

@@

Being a dynamically-typed language
with widespread use in a single domain,
R is a good target for a gradual type system.
A gradual type system permits the integration of
static and dynamic typing~\cite{siek:2006,tobin-hochstadt:2006},
and is capable of seamlessly mixing typed and untyped regions of code.

For statically-typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the type system precedes any code written
in the language.
Gradual type systems, by contrast, are often designed atop
an existing dynamically-typed language~\cite{tobin-hochstadt:2016}.
Many R programs already exist,
so a successful gradual type system must
work well with extant code.

We are not aware of any previous work that applies
visualization techniques for the purpose of programming language design.
However, there are existing publications on understanding
software through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces
to improve developer comprehension of large software systems.
Their tool applies pruning and summarization techniques to reduce trace size.
They focus on call topology,
describing the structure of function invocations,
rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
\typevis has elements of both,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information,
allowing the user to gain further understanding of the system
under examination.
This includes grouping methods that come from the same class---a
basic form of type data.

@@

The data set we used comes from
Turcotte et al.~\cite{turcotte:2020}
and consists of traces from \pkgNumAnalyzed packages,
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN are curated and satisfy
certain quality standards.
Out of the over $15,000$ packages
distributed on CRAN,
the subset of 412 packages considered here
all had code coverage above 65\%.
\todo{Confirm with Alexi this paragraph. Get proper BibTeX entry.}

Execution traces were recorded by running a dynamic analysis
over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of function type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, the function call {\tt !(FALSE)}
would be recorded as the type signature {\tt logical $\to$ logical},
meaning that the function {\tt !} took a logical value (i.e. a boolean)
and returned a logical value.
The type tags are assigned by the dynamic analysis
and refine the actual type tags at run-time.
For instance, the analysis
assigns the scalar tag {\tt double} to a singleton vector of double values
because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \typevis was designed for the analysis of types,
not all information in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

@@

We interviewed the primary researcher developing a gradual type system for R
to inventory relevant domain-specific tasks.
Additionally, we sought his and other researchers' feedback
throughout the creation of \typevis
to ensure it was adequately meeting requirements.
The domain-specific goals identified were:

@@

\typevis is an interactive web-based visualization
that permits real-time data exploration,
even with large data sets.
The default data set of R execution traces,
described in \S\ref{sec:data},
contains over a million rows.

\figref{fig:typevis} shows
the heart of \typevis---the type flow panel on the left
and the function filtering panel on the right.
Type signatures shown on the left reflect
the subset of the data selected on the right.
By default, if no subset is selected, the type flow
panel shows several of the most frequent function type signatures
in the entire data set.
%in its initial configuration.
Various aspects of the system,
such as the number of type signatures shown at once,
can be tweaked by the user if desired;
for this, a separate \settingsTab tab is provided.

@@

The right-hand panel of \figref{fig:typevis}
contains three elements: a search bar,
a package treemap,
and a function treemap.
All three components are linked.
Selecting a package
in the top treemap causes the bottom treemap
to show only functions defined in that package;
additionally, the type flow panel gets updated
with the most frequent function type signatures in the package.
Selecting a function in the bottom treemap
makes the type flow panel show only signatures
of that function.
Entering a function name in the search bar
updates the type flow panel immediately.

These different components support at least two modes
of search: quick function lookup and package exploration.
If a user has a specific function in mind,
they can type its name into the search bar and see its
type signatures straight away;
the search bar supports autocomplete that is
able to select one function among thousands.
If, on the other hand, one wants to browse
all available packages and functions,
the treemap view is more applicable.
The area of a block in the function treemap (package treemap)
conveys the log-scaled
frequency of calls to that function
(to all functions defined in that package).
%Area in the bottom treemap conveys the log-scaled
%frequency of function calls,
%and correspondingly the top treemap
%is scaled based on invocations
%of functions defined in that package.
Log-scaling and pagination keep
elements of the treemaps legible,
even when absolute size differences are significant.

The data set can additionally be filtered by analyzed package,
in other words, the package containing the call site of a function, not the definition site.
As mentioned in \S\ref{sec:data}, the data set contains information
from \pkgNumAnalyzed analyzed packages, which are all taken into account
by default. However, a user can limit the set of analyzed packages,
using a search bar in \settingsTab.

Although treemaps are often used for displaying
hierarchical data~\cite{shneiderman:1992},
in \typevis they have a flat structure.
The packages in our data are not deeply structured,
so the treemaps in \figref{fig:typevis} are only one level deep.
However, in many programming languages, like Java,
package systems contain deeply nested namespace hierarchies.
The treemap idiom is appropriate for navigation of such
hierarchies, and
%However, if the data set recorded deeply nested packages,
\typevis could be modified to handle that.

By default, \typevis allows only one function to be selected at a time,
but it can be configured for multiple function selection.
In the example from \figref{fig:typevis},
the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are both selected.
These functions are rendered simultaneously in the type flow
panel, allowing the user to make comparisons between them.
However, if the user wants to explore different functions rather than
compare them, the default one-function mode is more appropriate.

@@

Prominently featured in the center of \figref{fig:typevis}
is a parallel sets diagram encoding the type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara et al.~\cite{kosara:2006},
are ideally suited for type signatures.
Each data point can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n, \tau_r)$
corresponding to the function name,
argument types,
and return type.
Every component of that tuple is categorical,
and we are interested in visualizing the frequencies of these points.
This is the exact use-case for parallel sets.

Flows begin at a node labeling a function,
curve across each argument type in order,
and terminate at the return type.
Widths are proportional to how many times a function
is called with that signature during analysis.
Flow segments are filled with a hue-varying color scale,
determined by the type where the segment ends.
Beneath each node label is an approximate count of either:
how many times the function was invoked if the node represents a function,
or how often a value of that type was recorded during analysis if the node represents a type.

In \figref{fig:typevis},
the {\tt length.unit} function is called with
two different type signatures at run-time.
Most frequently, the function is used with the
signature {\tt double $\to$ integer},
but a substantial fraction of the time it is
used as {\tt double[] $\to$ integer}.
Here, {\tt double[]} stands for an array of
doubles with arbitrary dimension.

Interactivity significantly enhances the
usability of the parallel sets diagram.
Hovering over a flow will highlight the
entire path of the flow
and will fade out all other functions,
as shown in \figref{fig:typevis}.
When displaying many flows at the same time,
highlighting becomes critical
for user comprehension.
Additional quantitative information
about the flow is supplied on-demand while hovering.
Clicking a flow will focus on that function
only, filtering away all others;
double clicking outside flows will bring back the previous view.
The number of flows shown at a time is limited,
and will paginate if that threshold is exceeded.

Flow layout is a necessary consideration
for a legible visualization~\cite{sugiyama:1981}.
Figure~\ref{fig:decross} exemplifies what
happens if flow layout is not adequately computed.
The top layout was generated without any
flow decrossing algorithm,
while the bottom layout applies optimal decrossing.
Even with highlighting,
the top visualization is almost impossible to understand
due to excessive flow overlap.
To maintain real-time performance,
we use two different decrossing methods.
First, the system attempts to construct
a layout with the globally optimal solution,
minimizing the amount of flow crossing by
solving a mixed-integer linear program~\cite{ebner:2005}.
Unfortunately, for even moderately
sized graphs this can take too long to compute.
Therefore, after about $1$ second,
the optimal algorithm is terminated
if it has not finished, and the final layout is computed
by a faster
algorithm that only minimizes local crossing between layers.
%computes the layout.
While the resulting layout might not be ideal,
it is much better than the one without decrossing at all.

@@

Since the amount of data a user may query is large,
the application is split into a frontend and a backend.
The backend is responsible for storing the data
and fetching required information from the underlying database.
The frontend serves as an interactive interface to the data
and provides the visualization.
This avoids excessive memory consumption and speeds up computations,
as the client queries data from the server on a need-to-know basis.

@@

Designed with a specific use case in mind,
\typevis suffers from some shortcomings
and does not fully address the entire range of tasks
a language designer may need to perform.

In particular, \typevis does not allow for exploring 
type information at differing granularities.
R supports a range of data types
including ones containing dimensionality information.
For example, an R array may have type {\tt integer[3]}
indicating that it is an array of length $3$.
Every such type will get compressed down %into a single array of
to a simpler array
type {\tt integer[]}.
Reducing the number of types can make the visualization
more understandable, at the expensive of losing
precision in type information.
Ideally, a user would be able to interactively tune
the level of granularity based on their interest.
Such a feature would be especially helpful for R,
because values can come attached with lots of relevant
data such as the dynamic dispatch method
or user-specified metadata.

@@

We have presented \typevis, an interactive visualization
of run-time type information
that supports real-time analysis and exploration.
The tool was instantiated with a massive data set of
execution traces from a corpus of popular R packages.

\typevis provides facilities for filtering, searching,
understanding, and comparing type information.
Type signatures of function calls
are represented as flows traversing
each argument type and return type in sequence.
Quick function lookup is supported by a search bar,
with package and function treemaps facilitating
function exploration.
Various interactions provide additional details and
assist users with comprehension and navigation.

Our system is aimed at elucidating how R programmers
use the language in the field.
Specifically, the data and visualization will be used to
inform the design of a future gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any %dynamically-typed language.
language that supports run-time type information.
The visualization could yield insights
useful for purposes other than gradual typing,
and beyond what we have considered.

We hope that, in the future, programming language designers
will collect and use empirical data about their languages
more extensively.
Such data could be used to make informed decisions about design,
where the cost of making a mistake is high.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
\typevis is a first step in this direction.
